Mixed language processing increases cross-language phonetic transfer in Bengali-English bilinguals

### 2022-10-01 draft_BLC/BLC_AM-ID_manuscript.tex
table for participant info

@@

\newpage
\begin{myparindent}{0pt}
Abstract:\\
This study investigates the phonetic outcome of mixed-language speech in Bengali and Indian English, towards understanding cross-language interaction in highly proficient bilinguals.  We compare spectral properties of L2 vowels \nt{\ae} (common to L1, L2) and \nt{2} (absent in L1) in unilingual vs. mixed productions. Results reveal asymmetrical shifts in both vowels, towards related L1 categories. We interpret this as a temporary increase in cross-language phonetic interaction during mixed-language use, leading to a shift towards L1 norms, evidencing transfer effects on L2 vowels for the first time. We elicit mixed productions through two tasks: cued picture-naming and code-switching, to assess if experimental paradigm independently influences the behavior under study. Results reveal parallel patterns, but small magnitude differences, across paradigms. We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction, postulated discursive factors in code-switching, and the issue of comparability between paradigms in transfer studies. \\

@@

The present study examines phonetic transfer between Bengali and English in a group of highly proficient bilingual speakers in India. We measure spectral properties, F1 and F2, of two English vowels to ascertain if L1 influence on L2 increases during mixed-language use, relative to a participant's unilingual baseline production of L2. Mixed-language data is elicited in two switching paradigms: cued picture-naming and code-switching. We compare these results to assess if differences between the paradigms independently influence the outcome of phonetic interaction. The results demonstrate a time-invariant effect of transfer on L2 vowels during mixed-language production, and parallel pattern but different magnitude of transfer between the two paradigms. We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction and the role of connected speech in introducing discursive factors to transfer studies.

@@

\cite{grosjean1998studying} suggests that various aspects of the communicative setting, including exposure to (spoken or written) stimuli in multiple languages and awareness of the interlocutor's being bilingual, could trigger a bilingual language mode. The participants in the present study are in an environment which largely contains mixed-language input, multilingual interlocutors, and no stable ambient language. Therefore we expect that phonetic transfer during everyday language use, if any, takes place in a bilingual mode.  To preserve ecological validity, we elicited both unilingual and switched utterances in a bilingual language mode. Any observed differences in this paradigm would arguably result from interaction during online processing of sounds while switching between languages.

@@

In the present study, our research question concerns whether L2 vowels are produced differently in unilingual vs mixed conditions. Previous literature on phonetic transfer suggests that such shifts are not random, but rather directional and targeted. Although our primary research question merely concerns whether productions in unilingual vs mixed conditions differ, we wanted to understand how the direction of shift compares to corresponding Bengali categories. To estimate average baseline differences in vowel quality between the relevant English and Bengali categories, we processed and used existing speech data from a freely available Bengali corpus, SHRUTI \citep{shruticorpus}. The procedure is described below:

@@

The SHRUTI corpus is the largest free corpus of Western Bengali speech that we are aware of. It comprises of 21.64 hours of read speech, containing 7383 unique sentences (22012 unique words) from Bengali newspapers in different domains, covering the most frequently used words of the language. The speakers are 34 adults (age range 20-40, 75\% male, 25\% female) who have grown up in the Indian state of West Bengal, and speak standard colloquial Bengali. All of the speakers have university-level education or higher. Given these parallels in linguistic, regional, and socio-economic background, we believe that Bengali speech from this corpus is representative of the average target Bengali vowel categories that are of interest in this study.
To estimate average vowel positions, we extracted formant values from syllable-initial productions of \nt{\ae} and \nt{a:}, at 55\% into the vowel, to get estimates from the steady state, which should most closely approximate the production target. Next, we compared these to our participants' productions in the unilingual English condition to understand the relative positions of these categories. This is visualized in figure \ref{vowels_unilingual_ie_bengali}.

@@

\subsection*{Research questions and hypotheses}\label{questions_and_hypotheses}
\begin{itemize}
	\item \textbf{RQ1---L2 vowels}: Are L2 vowel categories of proficient bilingual speakers affected by dynamic phonetic transfer during mixed-language production?
	\item \textbf{H1---L2 vowels}: We expect to see differences in the spectral properties of L2 vowels in switched vs non-switched tokens
	\item \textbf{RQ2---Asymmetry}: Do patterns of transfer differ between the two vowel categories?
	\item \textbf{H2---Asymmetry}: Based on existing literature evidencing targeted and phonologically-constrained shifts, and comparisons with Bengali vowels from \cite{dutta2021}, we expect \nt{2} to be lowered and fronted in the mixed condition. \nt{\ae} is expected to be fronted. We expect the extent of shift in \nt{\ae} to be lesser than \nt{2}, as it is in a part of the IE vowel space that has greater vowel density.
	\item \textbf{RQ3---Paradigm}: Is there an observable difference between the extent of transfer in a cued picture-naming vs. a code-switching paradigm? 
	\item \textbf{H3---Paradigm}: While code-switching as a source of discursive factors that are likely to affect transfer behavior, this has not been systematically studied. Existing literature does not allow us to predict precisely what form this effect might take (c.f. discussion in Section \ref{paradigms}). Thus, this question is exploratory, and meant to serve as a starting point for more within-study comparisons 
\end{itemize}

@@

\subsection{Participants} \label{participants}
Participants were recruited through an open call circulated on social media at the EFL University campus in Hyderabad. Interested volunteers filled a brief language background questionnaire. These responses were used to screen participants, in the manner described below. 
As noted in section \ref{bengali_english_in_india}, India has an extremely multilingual an linguistically diverse population. The university campus comprises of students from different parts of the country and world, with varying L1s and different varieties of English. Our aim in screening participants was to minimize between-participant random variability in the study, while maintaining ecological validity considering the actual day-to-day settings in which transfer behaviors likely take place. We chose two metrics to do this: (i) where the participant grew up; and (ii) amount of formal education in Bengali. The first ensures that the variety of Bengali that individuals are exposed to in their surroundings is similar across participants. This would mean that the target representations for Bengali vowels would be expected to be similar. This was especially important because we were not eliciting unilingual L1 Bengali data in this study, and were using an existing data source to estimate the position of Bengali vowels in the formant space. Based on previous literature (c.f. section \ref{asymmetry between sounds}), we expect that L2 vowels shift towards related L1 categories during mixed language utterances. Thus, we wanted to ensure that any between-participant asymmetries in the pattern of shift reflected individual differences in transfer, rather than different acoustic targets. Another reason to control for location was that it is a fairly reliable predictor for language(s) of education, owing to how the education system functions in the country (we elaborate on this below).

The second metric is Medium of Instruction (MoI) in school. Most schools in India follow the Three Language Policy, wherein formal education is provided in English, Hindi, and a regional language. (The idea and implementation of this policy, and its implications for linguistic minorities, has been a matter of much public debate as well as scholarly work; for discussions see, e.g. \cite{tollefson2014language, jhingran2009hundreds, khubchandani1997language, mohanty2009multilingual, ramanathan2005rethinking}). Sustained formal education in a language is a reliable predictor of LSRW (listening, speaking, reading, writing) skills. Since this study uses written stimuli, we wanted to ensure that all participants were sufficiently comfortable with reading and speaking both Bengali and English, to avoid confounds and uncontrolled variability due to reading/task effort.

With these considerations in mind, we only recruited participants who had grown up in the state of West Bengal and lived there for the majority of their lives, had parents who spoke Bengali, and had received at least 5 years of formal education in Bengali (all respondents had received 5+ years of education in English and were pursuing university degrees that are taught in English), and self-reported proficiency in speaking, understanding, and reading Bengali.

A total of 28 bilingual speakers of Bengali and English (18 female) responded to the recruitment call. Using the inclusion criteria reported above, 10 volunteers (5 female, 5 male, age range 19 to 28) were invited to participate in a speech production study, and were compensated for their time. All participants reported normal hearing, and normal or corrected-to-normal vision. After completion of the study, we followed up with a detailed language background survey at a later date to learn more about the language profiles of the participants. This survey was a modified and consolidated version of three popular bilingualism profiling tools: the Bilingual Language Profile (BLP) \citep{blp}, the Language Experience and Proficiency Questionnaire (LEAP-Q) \citep{leap-q} which incorporates language history and self-assessed proficiency, and the Bilingual Switching Questionnaire (BSWQ) \citep{language_switching_questionnaire}. Together, these incorporate questions about language history, usage, attitudes, and switching experience. The complete questionnaire, along with responses, is available in the ``Data" component of the OSF repository for this project (c.f. section \ref{data_availability}).

In addition to Bengali and English, all participants reported knowing other languages, the most common being Hindi. Since this situation is representative of the present population and our study concerns only Bengali-English interaction, we only focused on these languages in the survey.
The average age of acquisition was 2.6 for English (range 0 i.e. since birth to 6) and 0.25 (range 0 to 2) for Bengali. Thus, all participants can be described as simultaneous bilinguals, having acquired both English and Bengali at a young age. They report being comfortable using both languages at a young age, with the average being 7.5 for English and 0.3 for Bengali (on a likert scale rating where 0=for as long as I can remember, 5=since I was 5 years old, and so on). As mentioned earlier, all participants had attended schools where Bengali and English were both used as Medium of Instruction (MoI). On average, participants had had more than 16 years (range 7 to 20+) of formal education in English and 11.8 years (range 8 to 15) in Bengali at the time of the study. As described in the inclusion criteria, all participants reported having spent over 15 years in a Bengali-speaking region. At the time of recording, they were living in or around the EFL University campus in Hyderabad, which has a linguistically diverse population. In terms of usage, participants show a range of language usage patterns. Almost 100\% of the participants report exclusively using Bengali with little to no English while interacting with family, and English at school/the workplace. However, language use with friends varies greatly, likely reflecting the language composition of the friend circle. 
The questionnaire asked participants to rate their proficiency in each language. All participants rated themselves as highly proficient in speaking and understanding both Bengali and English. While self-ratings for reading and writing English were at ceiling, ratings for Bengali literacy (reading and writing) showed some variation, with an average of 4.87 (on a 0-6 likert scale, range 3 to 6) for reading and 4.25 (range 3 to 6) for writing. This is a common situation for many educated sections of society, since institutions of higher education beyond the school level largely use English (for discussions about the language politics of higher education in India, see, for example, \cite{mohanty2009multilingual}). 
The language attitude section of the questionnaire reveals interesting insights about the linguistic landscape of the population, and how people experience language in such a setting. When asked how much ``like themselves" participants feel while using each language, response were at ceiling for both languages. Similarly, most participants identified equally strongly with both an ``English-speaking" and a ``Bengali-speaking" culture, possibly because for most people, such linguistic-cultural identities are not clearly delineated and not exclusive of one another. 
These responses shed light on how multiple languages fit seamlessly into different facets of daily life to create a ``continuous" linguistic experience (c.f. ``plurality" as defined by \cite{khubchandani2021plural}). In such a scenario, it is not surprising that language-switching is common, expected, as well as used in creative ways with communicative purpose. Responses to the section on language switching demonstrate this. On average, participants do not report being often unable to recall words in the target language while speaking Bengali (average rating 2.5 on a 1-6 likert scale where 1=never and 6=always) or English (average rating 2.3), indicating that switching is not strictly necessary for filling lexical gaps. Indeed, participants do not report switching between languages without awareness (1.8) or control (1.25). In spite of this, participants report a proclivity to code-switch (rating 3.5). Importantly, a large number of participants report volition while switching languages (4.12), and many report this to be situationally motivated (3.12). These responses indicate that participants are highly aware of their switching habits and move between languages for communicative purposes. This fact, combined with the linguistic situation described here and in section \ref{bengali_english_in_india}, are important to bear in mind while interpreting the results of a transfer study in the present population.

Given our inclusion criteria, we were able to recruit 10 participants for this study. Although existing studies in the field have used comparable participant numbers (e.g ), we acknowledge that this is a small sample size. As detailed above, even after a recruitment call that already controlled for language background and education level (recruiting Bengali-English bilinguals in a university campus), we were able to include only 10 out of 28 volunteers (35.7\%) in the final study. Note that the criteria used here establish only a minimal level of between-participant similarity in exposure, dominance, habits, and proficiency, such as is extremely common in most comparable studies. This demonstrates the degree of linguistic variability in the present population. Norms for experimental research on bilingualism have largely been established in a Western context. In settings where linguistic variability at the population level is much greater than these canonically studied groups, establishing comparable levels of experimental control requires significantly greater resources. Given that small sample sizes are likely to have low statistical power, it is fruitful to treat these findings as preliminary. Replication with larger sample sizes will help to confirm the observed effects, and provide much-needed data diversity (c.f. discussion in section \ref{introduction}). A more detailed discussion of power analyses in bilingualism research is included as supplementary material on OSF (c.f. section \ref{data_availability}). We hope that this can facilitate more conversation around establishing norms for power and sample size in our field.

@@

\subsection{Procedure} \label{procedure}
Participants were recorded individually in a sound attenuated room. %, using a ... microphone type/description?
Productions were recorded in Praat \citep{boersma2016praat} at a 44.1K sampling rate. Each participant completed four iterations each of two tasks--- cued picture-naming and code-switching. Utterances in the two conditions (unilingual and mixed) were randomly interspersed within each iteration. In the unilingual condition, the target word was preceded by another English word in the utterance. In the mixed condition, the target was preceded by a Bengali word and thus produced as a switch from Bengali. Every target word appeared twice during a test block; once in a unilingual and once in a mixed condition. Participants alternated between the two tasks, and were allowed to take a short break after each test block. The tasks are described below:

\begin{enumerate}[]
	\item Cued picture-naming: Participants were presented with slides in the following sequence:
	\begin{itemize}
		\item Language cue: a word in either English or Bengali orthography
		\item A picture displayed for 50 ms; named by the participant in either English or Bengali, depending on the language cue. The picture of a fish was used in all trials, prompting the participant to say either \nt{fIS} or \nt{ma: \textteshlig \super{h}}, depending on the language cue
		\item Target: an English word; read out by the participant
		\item Distracter math problem
	\end{itemize}
	Since all the target words were in English, the language cue was used to manipulate the preceding word, giving target words in non-switched (unilingual) and switched (mixed) utterances. The orthographies of English and Bengali are visually distinct, and were therefore used to cue language.

	\item Code-switching: Each slide contained one target word embedded in either an English (unilingual) or a Bengali (mixed; target produced as a switch from Bengali) carrier sentence which was read out by the participant. The carrier sentences are given below:
	\begin{itemize}
		\item Unilingual sentence: That is a yellow [\textit{target word}].
		\item Mixed-language sentence: \textipa{o-\:ta \ae k-\:ta kalo} [\textit{target word}]. 
	\end{itemize}

@@

Our study involved two IE vowels. Vowel category is expected to systematically affect formant frequencies (FFs). Thus, Vowel is a fixed effect in the models. The experimental treatment of interest is Context, i.e. whether the target item was presented in a unilingual (English) or mixed (Bengali) context. Our main research question (\bf{L2 vowels}) asks whether L2 vowel quality (formant frequencies) of proficient bilinguals is affected by the Context of utterance. We hypothesize that the direction of such shift in the mixed condition, if any, is towards a corresponding L1 category. Therefore, we expect the effect of Context on formant frequencies, i.e. the direction of shift, to be moderated by the vowel category (c.f. the discussion in Section \ref{asymmetry between sounds}, leading to our second research question \textbf{Asymmetry}). Thus, our main effect of interest is the interaction term Context*Vowel. This corresponds to the transfer pattern. Our third research question (\bf{Paradigm}) asks whether the language-mixing paradigm, or Task, affects the extent and direction of transfer. If this is the case, we expect Task to interact with the transfer term, giving a three way Task*Context*Vowel interaction. Since we measured FFs at five points in the vowel, Time is expected to affect the FF. However, the precise effect of formant dynamics is expected to depend on the vowel category. Thus, we expect the interaction of Vowel*Time to predict formant frequency. If transfer patterns change across time, then we should expect it to interact with the transfer term, giving a Context*Vowel*Time interaction.

@@

First, to test our first research question (\bf{L2 vowels}), we fitted a model with vowel*time, and vowel*context as fixed effects. The transfer term (vowel*context) is significant in this model (model coefficient/ unstandardized effect size ($\beta$)= -2.330e-01, standard error (SE)=  5.244e-02, t= -4.44, p=0.0002). Comparing this to a null model that lacks the transfer term shows that the full model is significantly better ($\chi^2(2) = 14.42, p=  0.0007$). To confirm whether the effect of context is moderated by vowel identity, we compared a second null model that contained Context as a fixed effect, but lacked the Context*Vowel interaction. The full model was better ($\chi^2(1) = 12.51, p= 0.0004$).

To test the hypothesis \textbf{Paradigm}, we fitted a model with vowel*time and task*vowel*context as fixed effects. This three-way interaction term is significant ($\beta$= -1.955e-01, SE=  5.453e-02, t= -3.59, p= 0.0003). We also find a significant effect of task on FFs ($\beta$=3.300e-01, SE=2.668e-02, t=12.371, p=2e-16). Comparing this model to a null model that lacks the fixed effect Task ($\chi^2(4) = 572.61, p=  2.2e-16$) shows that the full model is better. Because Task has an independent effect on F1, and adding interactions increases model parameters, we fitted a second null model with Task as a separate fixed effect without the interaction term to test whether the interaction term is needed. However, the full model was still a better fit for the data ($\chi^2(3) = 23.82, p=  2.722e-05$). Next, we tested two other null models to verify that the full model with the three-way interaction term is indeed the optimal model for the data, by replacing the three-way interaction term with: (i) vowel*context + task*context (to test if task only interacts with context, rather than the transfer term vowel*context). The full model is better ($\chi^2(2) = 23.28, p= 8.78e-06$); (ii) vowel*context + task*vowel (to test if task only interacts with vowel, rather than the transfer term vowel*context). The full model is better ($\chi^2(2) = 13.36, p= 0.001$). This shows that the three-way interaction is the best fit for the data. Because an independent effect of task on FFs is unexpected (c.f. model interpretation below), we wanted to examine if the effect might be local, i.e. decrease over time. We added a task*time interaction term to the full model, which is significant ($\beta$= -1.216e-01, SE=  1.347e-02, t= -9.028, p= 2e-16 ) and improves model fit ($\chi^2(1) = 81.22, p= 2.2e-16$).

@@

Vowel is not a significant predictor of F1. This is initially surprising, given that the phonological category is expected to systematically affect vowel quality. However, a boxplot visualization (Figure \ref{boxplot_F1}) shows that the participants in this study vary greatly both in the relative heights of the two IE categories \nt{2, \ae} and the extent of height difference between them. Note that these measures are only from the English (unilingual) condition, suggesting that this variability is a feature of the participants' IE vowel categories, independent of the effects of switching. Given that estimates for fixed effects are pooled across participants, this explains the lack of reliable differences between \nt{2} and \nt{\ae}.

The negative slope of the estimate for Context shows that F1 is lower, i.e. vowels are higher, in the unilingual (e) condition compared to the mixed (b) condition. Put more intuitively, vowels are lowered in the mixed condition. However, the lack significance indicates that using English words in mixed utterances does not lead to a uniform lowering/raising across vowels. Instead, the significant interaction between Vowel and Context signals that the extent and direction of shift is moderated by vowel category. There is an asymmetrical shift in the categories \nt{2} and \nt{\ae} in the mixed-language condition. Specifically, the Vowel*Context interaction has a negative slope, showing that the lowering is greater in \nt{2}. The lack of a significant Vowel*Context*Time interaction shows that transfer patterns do not change across time. This indicates a uniform scaling of formant trajectories in the mixed condition, suggesting that the underlying acoustic targets for these vowels are different during unilingual vs mixed utterances.

@@

\newpage
\section*{Tables}
\begin{table}[h]
	\caption{Fixed effects}
	%\centering
	\resizebox{\textwidth}{!}{%
		\begin{tabular}{@{}ll@{}}
			\toprule
			Variable name                                                 & Levels                                                            \\ \midrule
			Vowel                                                  & \nt{2}, \nt{\ae} \\
			Task (corresponding to the two paradigms)              & p (Picture: cued picture-naming), s (Sentence: code-switching)     \\
			Context (corresponding to the experimental conditions) & e (English: unilingual), b (Bengali: mixed)                       \\
			Time                                                   & continuous variable, range -1.6 to 1.6, scaled and centered\\
			 \bottomrule
		\end{tabular}%
	}
	\label{table variables}
\end{table}

>>>>

%\usepackage[noae]{Sweave} %to get bold text?

@@

\newpage
\begin{myparindent}{0pt}
Abstract:\\
This study investigates the phonetic outcome of mixed-language speech in Bengali and Indian English, towards understanding cross-language interaction in highly proficient bilinguals.  We compare spectral properties of L2 vowels \nt{\ae} (common to L1, L2) and \nt{2} (absent in L1) in code-switched (mixed) vs. non-switched productions. Results reveal asymmetrical shifts in both vowels during mixed productions, towards related L1 categories. We interpret this as a temporary increase in cross-language phonetic interaction during mixed-language use, leading to a shift towards L1 norms, evidencing transfer effects on L2 vowels. We elicit mixed productions through two tasks: cued picture-naming and code-switching, to assess if experimental paradigm independently influences the behavior under study. Results reveal parallel patterns, but small magnitude differences, across paradigms. We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction, postulated discursive factors in code-switching, and the issue of comparability between paradigms in transfer studies. \\

@@

The present study examines phonetic transfer between Bengali and English in a group of highly proficient bilingual speakers in India. We measure spectral properties, F1 and F2, of two English vowels to ascertain if L1 influence on L2 increases during mixed-language use, relative to a participant's baseline production of L2. Mixed-language data is elicited in two switching paradigms: cued picture-naming and code-switching. We compare these results to assess if differences between the paradigms independently influence the outcome of phonetic interaction. The results demonstrate a time-invariant effect of transfer on L2 vowels during mixed-language production, and parallel pattern but different magnitude of transfer between the two paradigms. We discuss these findings in light of recent proposals about asymmetries in short-term phonetic interaction and the role of connected speech in introducing discursive factors to transfer studies.

@@

\cite{grosjean1998studying} suggests that various aspects of the communicative setting, including exposure to (spoken or written) stimuli in multiple languages and awareness of the interlocutor's being bilingual, could trigger a bilingual language mode. The participants in the present study are in an environment which largely contains mixed-language input, multilingual interlocutors, and no stable ambient language. Therefore we expect that phonetic transfer during everyday language use, if any, takes place in a bilingual mode.  To preserve ecological validity, we elicited both switched and non-switched utterances in a bilingual language mode. Any observed differences in this paradigm would arguably result from interaction during online processing of sounds while switching between languages.

@@

In the present study, our research question concerns whether L2 vowels are produced differently in nonswitch vs switch conditions. Previous literature on phonetic transfer suggests that such shifts are not random, but rather directional and targeted. Although our primary research question merely concerns whether productions in the two conditions differ, we wanted to understand how the direction of shift compares to corresponding Bengali categories. To estimate average baseline differences in vowel quality between the relevant English and Bengali categories, we processed and used existing speech data from a freely available Bengali corpus, SHRUTI \citep{shruticorpus}. The procedure is described below:

@@

The SHRUTI corpus is the largest free corpus of Western Bengali speech that we are aware of. It comprises of 21.64 hours of read speech, containing 7383 unique sentences (22012 unique words) from Bengali newspapers in different domains, covering the most frequently used words of the language. The speakers are 34 adults (age range 20-40, 75\% male, 25\% female) who have grown up in the Indian state of West Bengal, and speak standard colloquial Bengali. All of the speakers have university-level education or higher. Given these parallels in linguistic, regional, and socio-economic background, we believe that Bengali speech from this corpus is representative of the average target Bengali vowel categories that are of interest in this study.
To estimate average vowel positions, we extracted formant values from syllable-initial productions of \nt{\ae} and \nt{a:}, at 55\% into the vowel, to get estimates from the steady state, which should most closely approximate the production target. Next, we compared these to our participants' productions in the non-switched English condition to understand the relative positions of these categories. This is visualized in figure \ref{vowels_unilingual_ie_bengali}.

@@

\subsection*{Research questions and hypotheses}\label{questions_and_hypotheses}
\begin{itemize}
	\item \textbf{RQ1---L2 vowels}: Are L2 vowel categories of proficient bilingual speakers affected by dynamic phonetic transfer during mixed-language production?
	\item \textbf{H1---L2 vowels}: We expect to see differences in the spectral properties of L2 vowels in switched vs non-switched tokens
	\item \textbf{RQ2---Asymmetry}: Do patterns of transfer differ between the two vowel categories?
	\item \textbf{H2---Asymmetry}: Since existing literature demonstrates targeted and phonologically-constrained shifts, we expect transfer effects to be category-specific. Based on the position of IE vowels compared to those of Bengali vowels from \cite{shruticorpus}, we expect \nt{2} to be lowered and fronted in the mixed condition. \nt{\ae} is expected to be fronted. We hypothesize that the extent of shift in \nt{\ae} to be lesser than \nt{2}, as it is in a part of the IE vowel space that has greater vowel density.
	\item \textbf{RQ3---Paradigm}: Is there an observable difference between the extent of transfer in a cued picture-naming vs. a code-switching paradigm? 
	\item \textbf{H3---Paradigm}: While researchers have discussed code-switching as a source of discursive factors likely to affect transfer behavior, this has not been systematically studied. Existing literature does not allow us to predict precisely what form this effect might take (c.f. discussion in Section \ref{paradigms}). Thus, this question is exploratory, and meant to serve as a starting point for more within-study comparisons. 
\end{itemize}

@@

\subsection{Participants} \label{participants}
Participants were recruited through an open call at the EFL University campus in Hyderabad. Interested volunteers completed a brief language background questionnaire which was used to screen participants as described below.

As noted in section \ref{bengali_english_in_india}, the population is multilingual and linguistically diverse. The university campus comprises of students from different parts of the country and world, with varying L1s and different varieties of English. Our aim in screening participants was to minimize between-participant random variability, while maintaining ecological validity considering the actual day-to-day settings in which transfer behaviors likely take place. We chose two metrics to do this: (i) where the participant grew up; and (ii) amount of formal education in Bengali. The first ensures that the variety of Bengali that individuals are exposed to in their surroundings is similar across participants. This would mean that the target representations for Bengali vowels would be expected to be similar. This was especially important because we were not eliciting unilingual L1 Bengali data in this study, and were using an existing data source to estimate the position of Bengali vowels in the formant space. Based on previous literature (c.f. section \ref{asymmetry between sounds}), we expect that L2 vowels shift towards related L1 categories during mixed language utterances. Thus, we wanted to ensure that any between-participant asymmetries in the pattern of shift reflected individual differences in transfer, rather than different acoustic targets. Another reason to control for location was that it is a fairly reliable predictor for language(s) of education(see Supplementary materials for details).

The second metric is Medium of Instruction (MoI) in school. Sustained formal education in a language is a reliable predictor of LSRW (listening, speaking, reading, writing) skills. Since this study uses written stimuli, we wanted to ensure that all participants were sufficiently comfortable with reading and speaking both Bengali and English, to avoid variability due to reading/task effort.

Given these considerations, we only recruited participants who had grown up in the state of West Bengal and lived there for the majority of their lives, had parents who both spoke Bengali, and had received at least 5 years of formal education in Bengali (all respondents had received 5+ years of education in English and were pursuing university degrees that are taught in English), and self-reported being proficient in speaking, understanding, and reading Bengali.

A total of 28 bilingual speakers of Bengali and English (18 female) responded to the recruitment call. Using the inclusion criteria reported above, 10 volunteers (5 female, 5 male, age range 19 to 28) were invited to participate in a speech production study, and were compensated for their time. After completion of the study, we followed up with a detailed language background survey at a later date to learn more about the language profiles of the participants. This survey was a modified and consolidated version of three popular bilingualism profiling tools: the Bilingual Language Profile (BLP) \citep{blp}, the Language Experience and Proficiency Questionnaire (LEAP-Q) \citep{leap-q} which incorporates language history and self-assessed proficiency, and the Bilingual Switching Questionnaire (BSWQ) \citep{language_switching_questionnaire}. Together, these incorporate questions about language history, usage, attitudes, and switching experience. \alert{The complete questionnaire, along with responses, is available in the ``Data" component of the OSF repository for this project (c.f. section \ref{data_availability}). --footnotes?}

Given our inclusion criteria, we were able to recruit 10 participants for this study. Although existing studies in the field have used comparable participant numbers, we acknowledge that this is a small sample size. As detailed above, even after a recruitment call that already controlled for language background and education level (recruiting Bengali-English bilinguals in a university campus), we were able to include only 10 out of 28 volunteers (35.7\%) in the final study. Note that the criteria used here establish only a minimal level of between-participant similarity in exposure, dominance, habits, and proficiency, such as is extremely common in most comparable studies. This demonstrates the degree of linguistic variability in the present population. Norms for experimental research on bilingualism have largely been established in a Western context. In settings where linguistic variability at the population level is much greater than these canonically studied groups, establishing comparable levels of experimental control requires significantly greater resources. Given that small sample sizes are likely to have low statistical power, replication with larger sample sizes are important to confirm the observed effects, and provide much-needed data diversity (c.f. discussion in section \ref{introduction}). A more detailed discussion of power analyses in bilingualism research is included as supplementary material on OSF (c.f. section \ref{data_availability}). We hope that this can facilitate more conversation around establishing norms for power and sample size in our field.

@@

\subsection{Procedure} \label{procedure}
Participants were recorded individually in a sound attenuated room. Productions were recorded in Praat \citep{boersma2016praat} at a 44.1K sampling rate. Each participant completed four iterations each of two tasks--- cued picture-naming and code-switching. Utterances in the two conditions (nonswitch and switch/mixed) were randomly interspersed within each iteration. In the nonswitch condition, the target word was preceded by another English word in the utterance. In the mixed condition, the target was preceded by a Bengali word and thus produced as a switch from Bengali. Every target word appeared twice during a test block; once in a nonswitch and once in a switch condition. Participants alternated between the two tasks, and were allowed to take a short break after each test block. The tasks are described below:

\begin{enumerate}[]
	\item Cued picture-naming: Participants were presented with slides in the following sequence:
	\begin{itemize}
		\item Language cue: a word in either English or Bengali orthography
		\item A picture displayed for 50 ms; named by the participant in either English or Bengali, depending on the language cue. The picture of a fish was used in all trials, prompting the participant to say either \nt{fIS} or \nt{ma: \textteshlig \super{h}}, depending on the language cue
		\item Target: an English word; read out by the participant
		\item Distracter math problem
	\end{itemize}
	Since all the target words were in English, the language cue was used to manipulate the preceding word, giving target words in non-switched and switched (mixed) utterances. The orthographies of English and Bengali are visually distinct, and were therefore used to cue language.

	\item Code-switching: Each slide contained one English target word embedded in either an English (nonswitch) or a Bengali (mixed; target produced as a switch from Bengali) carrier sentence which was read out by the participant. The carrier sentences are given below:
	\begin{itemize}
		\item Unilingual sentence: That is a yellow [\textit{target word}].
		\item Mixed-language sentence: \textipa{o-\:ta \ae k-\:ta kalo} [\textit{target word}]. 
	\end{itemize}

@@

Our study involved two IE vowels. Vowel category is expected to systematically affect formant frequencies (FFs). Thus, Vowel is a fixed effect in the models. The experimental treatment of interest is Context, i.e. whether the English target item was presented in an English (nonswitch) or Bengali (switch/mixed) context. Our main research question (\bf{L2 vowels}) asks whether L2 vowel quality (formant frequencies) of proficient bilinguals is affected by the Context of utterance. We hypothesize that the direction of such shift in the mixed condition, if any, is towards a corresponding L1 category. Therefore, we expect the effect of Context on formant frequencies, i.e. the direction of shift, to be moderated by the vowel category (c.f. the discussion in Section \ref{asymmetry between sounds}, leading to our second research question \textbf{Asymmetry}). Thus, the main effect of interest is the interaction term Context*Vowel. This corresponds to the transfer pattern. Our third research question (\bf{Paradigm}) asks whether the language-mixing paradigm, or Task, affects the extent and direction of transfer. If this is the case, we expect Task to interact with the transfer term, giving a three way Task*Context*Vowel interaction. Since we measured FFs at five points in the vowel, Time is expected to affect the FF. However, the precise effect of formant dynamics is expected to depend on the vowel category. Thus, we expect the interaction of Vowel*Time to predict formant frequency. If transfer patterns change across time, then we should expect it to interact with the transfer term, giving a Context*Vowel*Time interaction.

@@

First, to test our first and second research questions (\bf{L2 vowels and Asymmetry}), we fitted a model with vowel*time, and vowel*context as fixed effects. The transfer term (vowel*context) is significant in this model (model coefficient/ unstandardized effect size ($\beta$)= -2.330e-01, standard error (SE)=  5.244e-02, t= -4.44, p=0.0002). Comparing this to a null model that lacks the transfer term shows that the full model is significantly better ($\chi^2(2) = 14.42, p=  0.0007$). To confirm whether the effect of context is moderated by vowel identity, we compared a second null model that contained Context as a fixed effect, but lacked the Context*Vowel interaction. The full model was better ($\chi^2(1) = 12.51, p= 0.0004$).

To examine the third research question, \textbf{Paradigm}, we fitted a model with vowel*time and task*vowel*context as fixed effects. This three-way interaction term is significant ($\beta$= -1.955e-01, SE=  5.453e-02, t= -3.59, p= 0.0003). We also find a significant effect of task on FFs ($\beta$=3.300e-01, SE=2.668e-02, t=12.371, p=2e-16). Comparing this model to a null model that lacks the fixed effect Task ($\chi^2(4) = 572.61, p=  2.2e-16$) shows that the full model is better. Because Task has an independent effect on F1, and adding interactions increases model parameters, we fitted a second null model with Task as a separate fixed effect without the interaction term to test whether the interaction term is needed. However, the full model was still a better fit for the data ($\chi^2(3) = 23.82, p=  2.722e-05$). Next, we tested two other null models to verify that the full model with the three-way interaction term is indeed the optimal model for the data, by replacing the three-way interaction term with: (i) vowel*context + task*context (to test if task only interacts with context, rather than the transfer term vowel*context). The full model is better ($\chi^2(2) = 23.28, p= 8.78e-06$); (ii) vowel*context + task*vowel (to test if task only interacts with vowel, rather than the transfer term vowel*context). The full model is better ($\chi^2(2) = 13.36, p= 0.001$). This shows that the three-way interaction is the best fit for the data. Because an independent effect of task on FFs is unexpected (c.f. model interpretation below), we wanted to examine if the effect might be local, i.e. decrease over time. We added a task*time interaction term to the full model, which is significant ($\beta$= -1.216e-01, SE=  1.347e-02, t= -9.028, p= 2e-16 ) and improves model fit ($\chi^2(1) = 81.22, p= 2.2e-16$).

@@

Vowel is not a significant predictor of F1. This is initially surprising, given that the phonological category is expected to systematically affect vowel quality. However, a boxplot visualization (Figure \ref{boxplot_F1}) shows that the participants in this study vary greatly both in the relative heights of the two IE categories \nt{2, \ae} and the extent of height difference between them. Note that these measures are only from the nonswitch condition, suggesting that this variability is a feature of the participants' IE vowel categories, independent of the effects of switching. Given that estimates for fixed effects are pooled across participants, this explains the lack of reliable differences between \nt{2} and \nt{\ae}.

The negative slope of the estimate for Context shows that F1 is lower, i.e. vowels are higher, in the nonswitch (e) condition compared to the mixed (b) condition. Put more intuitively, vowels are lowered in the mixed condition. However, the lack significance indicates that using English words in mixed utterances does not lead to a uniform lowering/raising across vowels. Instead, the significant interaction between Vowel and Context signals that the extent and direction of shift is moderated by vowel category. There is an asymmetrical shift in the categories \nt{2} and \nt{\ae} in the mixed-language condition. Specifically, the Vowel*Context interaction has a negative slope, showing that the lowering is greater in \nt{2}. The lack of a significant Vowel*Context*Time interaction shows that transfer patterns do not change across time. This indicates a uniform scaling of formant trajectories in the mixed condition, suggesting that the underlying acoustic targets for these vowels are different during non-switched vs mixed utterances.

@@

\newpage
\section*{Tables}
